ODGI: understanding pangenome graphs

### 2021-11-01 manuscript/main.tex
thinner bandage layout; formatting

Pangenome graphs model alignments of many genomes.
With \textit{odgi untangle}, users can extract pairwise 1:many alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references)
This lets us decompose both homology and paralogy relationships by also untangling loopy motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
Being able to work with any sets of reference sequences lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
As an example, by untangling the graph we can study the variation that lies in the collapsed regions due to sequence repeats (as shown in Fig.~\ref{fig:odgi_viz}\textbf{f}).
Indeed, to obtain a more precise overview of the \textit{locus} in Fig.~\ref{fig:odgi_viz}\textbf{b-e},
we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.

@@

Pangenome graphs stand to become a ubiquitous tool in genomics~\citep{Eizenga_2020}. With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
%
Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs. ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs. Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}. %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.

>>>>

Pangenome graphs model alignments of many genomes.
With \textit{odgi untangle}, users can extract pairwise 1:many alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references)
This lets us decompose both homology and paralogy relationships by also untangling loopy motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
Being able to work with any sets of reference sequences lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
As an example, by untangling the graph we can study the variation that lies in the collapsed regions due to sequence repeats (as shown in Fig.~\ref{fig:odgi_viz}\textbf{f}).
Indeed, to obtain a more precise overview of the \textit{locus} in Fig.~\ref{fig:odgi_viz}\textbf{b-e},
we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them by using the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.

@@

Pangenome graphs stand to become a ubiquitous tool in genomics~\citep{Eizenga_2020}.
With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.

Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs.
Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome~\citep{pangenome}. %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.
